Space is the Place: Effects of Continuous Spatial Structure on Analysis of Population Genetic Data

### 2019-11-18 ms/review-responses.tex

\begin{point}{}
    Related to the first point, to what extent are the results qualitatively different to those that would be obtained in a stepping-stone model? My interpretation is that they are actually very similar, but I didn't see whether that was explicitly discussed. In some sense, it's still easier to do large simulations in a stepping-stone model so it would be nice to be reassured that that's still ok.
\end{point}

\reply{
    See above. It is certainly easier and faster, but comes with significant issues either when neighborhood size is lower than the population of a deme, or when demes are small enough that the sample size approaches local $N_e$, at least in reverse-time simulations. 
}

\begin{point}{}
    The source of equation \eqref{eqn:p_survival} is not obvious to me. I sort of see how it makes sense, but a little but more intuition or a brief derivation or an illuminating either in the main text or the supplement, would be helpful.
\end{point}

\reply{
    Additional discussion of the model is now provided in the appendix (see "Demographic model") \llname{dem_model_app}
}

@@

\begin{point}{}
    The authors use a scaling factor in equation (2) to counteract the increase in fitness of individuals at the edges. Can they provide a figure showing that this is the case. What does ``roughly'' mean on line 164. Perhaps a heatmap of the fitness of individuals across the grid with and without the scaling factor?
\end{point}

\reply{
Good suggestion -- we have added a supplemental figure to show the distribution of individual fitness across the landscape with and without our edge-scaling approach. (Note though that we found the effect subtle and hard to visualize well). 
}

@@

\begin{point}{}
    It would be helpful provide the figure showing that generating mutations during the forward simulations in SLIM is equivalent to applying mutations using msprime on pre-generated trees (line 185)? It sounds like this procedure would underestimate the variance in the number of mutations, since you remove the effect of random generation time. Is this effect small?
\end{point}

\reply{
    We have added a figure showing sample site-frequency-spectra generated from a subset of simulations run with SLiM mutations, and then using msprime to apply mutations to the same tree sequences with our generation-time scaling approach \ref{fig:slim_v_msp_sfs}. These approaches yield extremely similar spectra so we believe any error in the approximation is quite small. We also now show variation in simulation times with different mutation and tree sequence recording schemes in Figure \ref{fig:runtimes}, and included a brief discussion of this in \llname{runtimesxn}
}

\begin{point}{}
    Can the authors provide a bit more intuition behind the patterns of variation seen in generation time, census population size, and variance in the number of offspring with respect to neighborhood size seen in Figure 2? For example, it is not obvious to me why the census population size, for example, should decline systematically with respect to neighborhood size. Presumably this isn't just due to the local demographic stochasticity. Could the authors briefly interpret the observed patterns or cite appropriate literature?
\end{point}

\reply{
    We have added a brief discussion of the causes of the census size scaling to the relevant results section \llname{censusscaling}.
}

@@

\end{point}

\reply{
    Similarly here we monitored the population size over time while running SLiM. All simulations maintained asymptotic populations, and none of the runs we started ended because populations crashed.
}

\begin{point}{}
    Third, it is not clear what the relationship between the expected lifespan and the probability of survival is. The expected lifespan, L, is first defined as the inverse of the expected number of offspring produced by a parent. The authors also define the probability of survival of a given individual at a given point in space, $p_i$. Hence, the expected lifespan at a point in space (and time) is the mean of a geometric distribution with parameter $p_i$, i.e., $1/p_i$. Now, it is far from being obvious what the relationship between these two approaches for defining the expected lifespan actually is.
\end{point}

\reply{
   We have added additional discussion of this point to appendix A (see "Demographic Model") \llname{dem_model_app}
}

@@

\reply{
    This is a good point also raised by other reviewers. We have added an Appendix comparing our model to a stepping-stone, rather than the spatial-Labmda-Fleming-Viot, because we think these are the most familiar and widely used class of spatial models. See Appendix 1. We find that many features of our model are well approximated by stepping-stone models, and that for statistics like $\theta_{\pi}$ the stepping stone model results approach our continuous space model as the number of demes used to describe the landscape increases. 
}

\begin{point}{}
    Figure 2: I do not understand why the neighborhood size varies to the same extent in the random mating model as it does for the spatial model. For the random mating model, I would have expected the neighborhood size to be equal to the census size since all individuals have the same probability of being a parent of any given offspring. From lines 166->171, it is clear that the spatial model would converge to the random mating model when the mean parent-offspring distance tends to infinity only if we were to ignore the impact of range edges. I am thus wondering whether the variation of neighborhood size one observes in Fig 2 for the random mating model is just a consequence of border effects. If that is the case, then the authors should state it clearly and try to justify it from a biological perspective.
\end{point}

\reply{
    We have a added a brief discussion of the census size scaling to \llname{censusscaling}
}

@@

\reply{
    Thank you for this suggestion. We have slightly revised the introduction and hope it is now clearer; however since we want to cover a little history and motivation for our continuous model vs stepping-stone approaches in the intro it does have a different flow from the discussion.

}

\begin{point}{}
    I agree that most modern work describes structure as discrete populations connected by migration. However, some methods/studies have explicitly modeled spatial structure, e.g. especially in ecology or using methods like dadi (diffusion approximations). Highlighting some examples of previously identified structure not possible to infer without modeling geography would be helpful to contextualize this work.
\end{point}

\reply{
   We have expanded our citations of some of the relevant ecology literature \llname{ecolcites}, which we hope helps to contextualize the study better.
}

@@

\begin{point}{}
    There is some reference to spatial models using grids (e.g. Rousset 1997). Some additional discussion contextualizing more recent methods like EEMS that also construct demes and model migration through divergence between neighboring demes would be helpful and interesting.
\end{point}

\reply{
    Good point on EEMS. We have added the most recent EEMS paper to the citations on \llname{eemscitation}, and have added an Appendix giving a more thorough comparison with stepping-stone models. 
}

@@

\begin{point}{}
    Demographic modeling. Both approaches tested, stairwayplot and SMC++, are most sensitive to older demographic events, and consequently are very noisy and underestimate effect population sizes, especially in smaller neighborhood sizes. Models that consider haplotype structure are much better suited to this time period. It would be helpful to either 1) discuss the varying time sensitivities of different classes of demographic inference methods and how spatial patterns of genetic variation would influence these inferences, or 2) apply a method of this class (many options, e.g. DoRIS, IBDNe, Tracts, Globetrotter, etc) and show how it performs.
\end{point}

\reply{
    We now discuss haplotype methods in the relevant discussion section \llname{ibdnesxn}. However though these methods should be more accurate for recent events it is not clear that this will improve performance per se. The dips in recent inferred Ne from stairwayplot are not just prediction noise, but actually reflect an underlying genealogy in which terminal branches are shorter than expected from a constant-size random-mating population (see e.g. figure \ref{fig:sumstats}A and \ref{fig:ibs}). The interpretation error is that these short branches are generated by spatial structure rather than changes in population size over time -- a point also made in the \citep{Mazet2015} paper we discuss in the introduction and discussion. 
}

\begin{point}{}
    GWAS mixed models. To what extent can spatial signals (e.g. corner, patchy) be corrected with mixed models, e.g. with PCs and PC-adjusted GRM as in Conomos et al, 2016 using PC-AiR and PC-Relate)? Is patchiness related to dispersal? I'm curious how this relates to the predictive ability of GWAS phenotypes with some spatial association that may or may not be associated with environmental effects.
\end{point}

\reply{
    Good question -- we are also interested to know how mixed models perform here, but think that to properly test that we would want to change our design to generate phenotypes from simulated genotypes. This would allow us to evaluate false-negatives in addition to false-positives. This is important because, if mixed models do provide stronger control for stratification they are also likely to remove true signals of causal SNPs if those SNPs covary with spatial structure. We now point to these methods explicitly in the discussion \llname{mixedmodelcallout}, but think that incorporating that study here would make this paper too long. We also think the PC results are still quite relevant as the method is still seen in many studies. 
}

### 2019-11-18 ms/spaceness.tex
pass through PR edits, cleaned up response comments, added edge scaling fig

To avoid hard-to-evaluate approximations, we here used forward-time, individual-based simulations.
The question of what regulates real populations has a long history and many answers
\citep[e.g.,][]{lloyd1967crowding,antonovics1980ecological,crawley1990population},
but it is clear that populations must at some point have density-dependent feedback on population size,
or else they would face eventual extinction or explosion.
In the absence of unrealistic global population regulation,
this regulation must be local,
and there are many ways to arrange this \citep{bolker2003spatial}.
In our simulations,
we decrease each individual's probability of survival with increasing local population density,
which shifts reproductive output towards low-density regions 
(and prevents the unrealistic population clustering seen by \citet{Felsenstein1975},
see Supplemental \autoref{fig:slim_has_no_pain}).
Such models have been used extensively in ecological modeling
\citep{durrett1994importance,bolker1997using,law2003population,fournier2004microscopic,champer2019suppression}
but rarely in population genetics, \revpoint{AE}{2}
where to our knowledge implementations of continuous space models
before their availability through SLiM \citep{Haller2019}
have focused on a small number of genetic loci
\citep[e.g.,][]{slatkin1989comparison,barton2002neutral,robledoarnuncio2010isolation,rossine2014espao},
which limits the ability to investigate the impacts of continuous space on genome-wide genetic variation as is now routinely sampled from real organisms. 
By simulating chromosome-scale sequence alignments and complete population histories we are able to treat our simulations as real populations and replicate the sampling designs and analyses commonly conducted on real genomic data.

@@

An important step in creating any spatial model is dealing with range edges. 
Because local population density is used to model competition, edge or corner populations can be assigned artificially high fitness values 
because they lack neighbors within their interaction radius but outside the bounds of the simulation.
We approximate a decline in habitat suitability near edges by decreasing the probability of survival proportional to the square root of distance to edges in units of $\sigma$. 
The final probability of survival for individual $i$ is then
\begin{equation}
    s_{i} = p_{i} 
        \min(1,\sqrt{x_{i}/\sigma})
        \min(1,\sqrt{y_{i}/\sigma})\\
        \min(1,\sqrt{(W - x_{i})/\sigma})
        \min(1,\sqrt{(W - y_{i})/\sigma)}
\end{equation}
where $x_i$ and $y_i$ are the spatial coordinates of individual $i$, and $W$ is the width (and height) of the square habitat.
This buffer roughly counteracts the increase in fitness individuals close to the edge would otherwise have.

@@

There are a few additional things to note about Figure~\ref{fig:genparams}. 
First, all three quantities are non-monotone with neighborhood size. 
Census size largely declines as neighborhood size increases for both the spatial and random mating models. 
However, for spatial models this decline only begins for neighborhood size $\geq 10$. 
By a neighborhood sizes larger than 100, the spatial and random mating models are indistinguishable from one another, 
a sign that our simulations are performing as expected. 
Census sizes range from $\approx 14,000$ at low $\sigma$ in the random mating model to $\approx 10,000$ for both models when neighborhood sizes approach 1,000.

@@

Run times for our model scale approximately linearly with neighborhood size (Figure \ref{fig:runtimes}), with the lowest neighborhood sizes reaching 30N generations in around a day and those with neighborhood size approaching 1,000 requiring up to two weeks of computation. At small neighborhood sizes the overhead caused by processing mutations and simplifying tree sequences can add $\approx30\%$ to the total run time, but above neighborhood sizes of $\approx$ 100 the effects of neighborhood size dominate because our simulation requires pairwise calculations for sets of individuals within a $3\sigma$ radius. As currently implemented running simulations at neighborhood sizes $>1000$ to coalescence is likely impractical, though running these models for more limited timescales is possible. \llabel{runtimesxn}

%The fitness scaling procedure we use to avoid high fitness at range edges shrinks the suitable area by one $\sigma$ on all sides and likely explains part of the variation in census size, but scaling in offspring variance and the U-shaped distribution of census size in spatial models suggests additional processes at work. For spatial models census size initially rises as neighborhood sizes scale from 2 - 10, which seems to reflect an Allee effect \citep{Allee1949} in which some individuals are unable to find mates when the mate selection radius is very small.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Demographic model}
\label{apx:demographic_model}

@@

>>>>

To avoid hard-to-evaluate approximations, we here used forward-time, individual-based simulations.
The question of what regulates real populations has a long history and many answers
\citep[e.g.,][]{lloyd1967crowding,antonovics1980ecological,crawley1990population}, \llabel{ecolcites}
but it is clear that populations must at some point have density-dependent feedback on population size,
or else they would face eventual extinction or explosion.
In the absence of unrealistic global population regulation,
this regulation must be local,
and there are many ways to arrange this \citep{bolker2003spatial}.
In our simulations,
we decrease each individual's probability of survival with increasing local population density,
which shifts reproductive output towards low-density regions 
(and prevents the unrealistic population clustering seen by \citet{Felsenstein1975},
see Supplemental \autoref{fig:slim_has_no_pain}).
Such models have been used extensively in ecological modeling
\citep{durrett1994importance,bolker1997using,law2003population,fournier2004microscopic,champer2019suppression}
but rarely in population genetics, \revpoint{AE}{2}
where to our knowledge implementations of continuous space models
before their availability through SLiM \citep{Haller2019}
have focused on a small number of genetic loci
\citep[e.g.,][]{slatkin1989comparison,barton2002neutral,robledoarnuncio2010isolation,rossine2014espao},
which limits the ability to investigate the impacts of continuous space on genome-wide genetic variation as is now routinely sampled from real organisms. 
By simulating chromosome-scale sequence alignments and complete population histories we are able to treat our simulations as real populations and replicate the sampling designs and analyses commonly conducted on real genomic data.

@@

An important step in creating any spatial model is dealing with range edges. 
Because local population density is used to model competition, edge or corner populations can be assigned artificially high fitness values 
because they lack neighbors within their interaction radius but outside the bounds of the simulation.
We approximate a decline in habitat suitability near edges by decreasing the probability of survival proportional to the square root of distance to edges in units of $\sigma$. 
The final probability of survival for individual $i$ is then
\begin{equation}
    s_{i} = p_{i} 
        \min(1,\sqrt{x_{i}/\sigma})
        \min(1,\sqrt{y_{i}/\sigma})\\
        \min(1,\sqrt{(W - x_{i})/\sigma})
        \min(1,\sqrt{(W - y_{i})/\sigma)}
\end{equation}
where $x_i$ and $y_i$ are the spatial coordinates of individual $i$, and $W$ is the width (and height) of the square habitat.
This buffer roughly counteracts the increase in fitness individuals close to the edge would otherwise have, though the effect is relatively subtle \ref{fig:edge_scaling}.

@@

There are a few additional things to note about Figure~\ref{fig:genparams}. 
First, all three quantities are non-monotone with neighborhood size. 
Census size largely declines as neighborhood size increases for both the spatial and random mating models. 
However, for spatial models this decline only begins for neighborhood size $\geq 10$. 
By a neighborhood sizes larger than 100, the spatial and random mating models are indistinguishable from one another, 
a sign that our simulations are performing as expected. 
Census sizes range from $\approx 14,000$ at low $\sigma$ in the random mating model to $\approx 10,000$ for both models when neighborhood sizes approach 1,000. The scaling of census sizes in both random-mating and spatial models appears to be related to two consequences of the spatial competition function: the decline of fitness at range edges, which effectively reduces the habitable area by one $\sigma$ around the edge of the map and so results in a smaller habitable area at high $\sigma$ values; and variation in the equilibrium population density given varying competition radii. For spatial models we see an additional effect in which census size initially rises as neighborhood sizes scale from 2 - 10. This may reflect an Allee effect \citep{Allee1949} in which some individuals are unable to find mates when the mate selection radius is very small.  \llabel{censusscaling}

@@

Run times for our model scale approximately linearly with neighborhood size (Figure \ref{fig:runtimes}), with the lowest neighborhood sizes reaching 30N generations in around a day and those with neighborhood size approaching 1,000 requiring up to three weeks of computation. As currently implemented running simulations at neighborhood sizes $>1000$ to coalescence is likely impractical, though running these models for more limited timescales and then "completing" the simulation by starting a reverse-time simulation from the resulting tree sequence in msprime is possible.  \llabel{runtimesxn}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Demographic model}
\label{apx:demographic_model}
\llabel{dem_model_app}

@@

\begin{figure*}[p]
\centering
\includegraphics[width=0.6\textwidth]{figures/edge_scaling_comp.pdf}
\caption{Comparison of individual fitness across the landscape in simulations with (right) and without (left) a decline in fitness approaching range edges. Note the slight excess of high-fitness individuals at edges on the left, which is (partially) counteracted by the scaling procedure.}
\label{fig:edge_scaling} 
\end{figure*}
\FloatBarrier
